---
title: "How to Tell Someone's Age When All You Know Is Her Name"
author: "Emily Kothe"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    keep_md: yes
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
knitr::opts_chunk$set(fig.width=6, fig.height=6) 

library(ggplot2)
library(matrixStats)
library(dplyr)
library(babynames)
library(stringr)
library(lubridate)
library(Hmisc)

lifetables_us <- babynames::lifetables
```

This vignette is based on this FiveThirtyEight article 
[How to Tell Someone’s Age When All You Know Is Her Name](https://fivethirtyeight.com/features/how-to-tell-someones-age-when-all-you-know-is-her-name/#fn-1). 
That article demonstrated how you could use actuarial life tables and a database 
of baby names to estimate the age of living Americans with a given name.^[Like 
FiveThirtyEight we assume that people of the same sex die at the same rate 
regardless of their name - in practice will vary by a number of factors related 
to both name and life expectancy such as ethnicity. Because US lifetables are 
calculated in 10 year intervals we have interpolated values for death rates 
between intervals].

## Joseph

Like fivethirtyeight we start be examining the boys name "Joseph".

We filter the SSA name data to only include boys named Joseph and join it with 
the actuarial life table. Using linear interpolation and the life table 
data we estimate the proportion of Josephs born in any given year that are still 
alive today. 

```{r message=FALSE, warning = FALSE}
name_of_interest <- "Joseph"
sex_of_interest <- "M"

df <- lifetables_us %>% 
  filter(x == (lubridate::year(Sys.Date()) - year)) %>% 
  mutate(prop_still_alive = lx/100000) %>% 
  right_join(filter(babynames, 
                    name == name_of_interest, 
                    sex == sex_of_interest), 
             by= c("year", "sex")) %>% 
  mutate(prop_still_alive = case_when(
    year < 1900 ~ 0,  # The lifetables begin in 1910 but name data begins in 1900. 
                      # We assume no 118 year olds are still alive
    year == 2015 ~ 1, # For interpolation assume 100% of people from the last 
                      # year of the lifetable are still alive
    TRUE ~ prop_still_alive), 
  Name = case_when(
    is.na(name) ~ name_of_interest, 
    TRUE ~ name), 
  Count = case_when(
    is.na(n) ~ 0, 
    TRUE ~ as.numeric(n)), 
  interp = zoo::na.approx(prop_still_alive, 
                          na.rm = FALSE),
  alive_count = Count * interp)

med_age <- df %>% 
  summarise(wted_median = weightedMedian(x, 
                                         w = alive_count, 
                                         na.rm = TRUE))
peak <- df %>% 
  filter(n == max(df$Count))

```

We can then generate the plot below which shows the number of Josephs born each 
year and the number still estimated to be alive today. 

```{r}
ggplot(data = df, aes(x = year)) +
  geom_line(aes(y = Count), size = 1)+
  geom_col(aes(y = alive_count), fill="cornflowerblue") +
  geom_vline(data = peak, xintercept = peak$year, size = 1, linetype = 2)+
  scale_x_continuous(limits = c(1900, 2016)) +
  coord_cartesian(expand = FALSE) +
  theme_classic() + 
  labs(title = "Age Distribution of American Boys Named Joseph", 
       caption = "Source: Social Security Adminstration")
```

As we can see, `r name_of_interest` peaked in `r peak$year` when `r peak$n` 
`r name_of_interest`s were born. Based on the actuarial data, 
25583 `r name_of_interest`'s from `r peak$year` 
would still be alive in `r lubridate::year(Sys.Date())`. Overall, 
`r round(sum(df$alive_count, na.rm=TRUE)/sum(df$n, na.rm=TRUE)*100,0)`% of 
`r name_of_interest`s born between 1900 and 2017 would still be alive in 
`r lubridate::year(Sys.Date())`. The median age of an `r name_of_interest` 
still alive, would be `r round(med_age$wted_median, 0)` years.


## Brittany

Because the name Joseph has been enduringly popular, we can't make strong 
inferences about a person's age if all we know is that they are named Joseph. 
On the other hand, we can make much more confident guesses about a woman's age 
if we know she is named Brittany. 
 
```{r message=FALSE, warning = FALSE, echo=FALSE}
name_of_interest <- "Brittany"
sex_of_interest <- "F"

df <- lifetables_us %>% 
  filter(x == (lubridate::year(Sys.Date()) - year)) %>% 
  mutate(prop_still_alive = lx/100000) %>% 
  right_join(filter(babynames, 
                    name == name_of_interest, 
                    sex == sex_of_interest), 
             by= c("year", "sex")) %>% 
  mutate(prop_still_alive = case_when(
    year < 1900 ~ 0, 
    year == 2015 ~ 1,
    TRUE ~ prop_still_alive), 
  Name = case_when(
    is.na(name) ~ name_of_interest, 
    TRUE ~ name), 
  Count = case_when(
    is.na(n) ~ 0, 
    TRUE ~ as.numeric(n)), 
  interp = zoo::na.approx(prop_still_alive, 
                          na.rm = FALSE),
  alive_count = Count * interp)

med_age <- df %>% 
  summarise(wted_median = weightedMedian(x, 
                                         w = alive_count, 
                                         na.rm = TRUE))
peak <- df %>% 
  filter(n == max(df$Count))

ggplot(data = df, aes(x = year)) +
  geom_line(aes(y = Count), size = 1)+
  geom_col(aes(y = alive_count), fill="cornflowerblue") +
  geom_vline(data = peak, xintercept = peak$year, size = 1, linetype = 2)+
  scale_x_continuous(limits = c(1900, 2016)) +
  coord_cartesian(expand = FALSE) +
  theme_classic() + 
  labs(title = "Age Distribution of American Girls Named Brittany", 
       caption = "Source: Social Security Adminstration")
```

`r name_of_interest` peaked in `r peak$year` when `r max(df$n)` 
`r name_of_interest`s were born. Based on the actuarial data, 
37100 `r name_of_interest`'s from `r peak$year` 
would still be alive in `r lubridate::year(Sys.Date())`. Overall, 
`r round(sum(df$alive_count, na.rm=TRUE)/sum(df$n, na.rm=TRUE)*100,0)`% of 
`r name_of_interest`s born between 1900 and 2017 would still be alive in 
`r lubridate::year(Sys.Date())`. The median age of an `r name_of_interest` 
still alive, would be `r round(med_age$wted_median, 0)` years.


## Unisex Names 

It's also interesting to look at names that have become unisex over time. Here 
we use Harper as an example.

```{r message=FALSE, warning = FALSE, echo=FALSE}
name_of_interest <- "Harper"
sex_of_interest <- "F"

df <- lifetables_us %>% 
  filter(x == (lubridate::year(Sys.Date()) - year)) %>% 
  mutate(prop_still_alive = lx/100000) %>% 
  right_join(filter(babynames, 
                    name == name_of_interest, 
                    sex == sex_of_interest), 
             by= c("year", "sex")) %>% 
  mutate(prop_still_alive = case_when(
    year < 1900 ~ 0, 
    year == 2015 ~ 1,
    TRUE ~ prop_still_alive), 
  Name = case_when(
    is.na(name) ~ name_of_interest, 
    TRUE ~ name), 
  Count = case_when(
    is.na(n) ~ 0, 
    TRUE ~ as.numeric(n)), 
  interp = zoo::na.approx(prop_still_alive, 
                          na.rm = FALSE),
  alive_count = Count * interp)

med_age <- df %>% 
  summarise(wted_median = weightedMedian(x, 
                                         w = alive_count, 
                                         na.rm = TRUE))
peak <- df %>% 
  filter(n == max(df$Count))

ggplot(data = df, aes(x = year)) +
  geom_line(aes(y = Count), size = 1)+
  geom_col(aes(y = alive_count), fill="cornflowerblue") +
  geom_vline(data = peak, xintercept = peak$year, size = 1, linetype = 2)+
  scale_x_continuous(limits = c(1900, 2016)) +
  coord_cartesian(expand = FALSE) +
  theme_classic() + 
  labs(title = "Age Distribution of American Girls Named Harper", 
       caption = "Source: Social Security Adminstration")
```

For girls, `r name_of_interest` peaked in `r peak$year` when `r max(df$n)` 
`r name_of_interest`s were born. Based on the actuarial data, 
10241 female `r name_of_interest`'s from `r peak$year` 
would still be alive in `r lubridate::year(Sys.Date())`. Overall, 
`r round(sum(df$alive_count, na.rm=TRUE)/sum(df$n, na.rm=TRUE)*100,0)`% of a girl 
named `r name_of_interest` born between 1900 and 2017 would still be alive in 
`r lubridate::year(Sys.Date())`. The median age of a girl named 
`r name_of_interest` still alive, would be `r round(med_age$wted_median, 0)` years.

```{r message=FALSE, warning = FALSE, echo=FALSE}
name_of_interest <- "Harper"
sex_of_interest <- "M"

df <- lifetables_us %>% 
  filter(x == (lubridate::year(Sys.Date()) - year)) %>% 
  mutate(prop_still_alive = lx/100000) %>% 
  right_join(filter(babynames, 
                    name == name_of_interest, 
                    sex == sex_of_interest), 
             by= c("year", "sex")) %>% 
  mutate(prop_still_alive = case_when(
    year < 1900 ~ 0, 
    year == 2015 ~ 1,
    TRUE ~ prop_still_alive), 
  Name = case_when(
    is.na(name) ~ name_of_interest, 
    TRUE ~ name), 
  Count = case_when(
    is.na(n) ~ 0, 
    TRUE ~ as.numeric(n)), 
  interp = zoo::na.approx(prop_still_alive, 
                          na.rm = FALSE),
  alive_count = Count * interp)

med_age <- df %>% 
  summarise(wted_median = weightedMedian(x, 
                                         w = alive_count, 
                                         na.rm = TRUE))
peak <- df %>% 
  filter(n == max(df$Count))

ggplot(data = df, aes(x = year)) +
  geom_line(aes(y = Count), size = 1)+
  geom_col(aes(y = alive_count), fill="cornflowerblue") +
  geom_vline(data = peak, xintercept = peak$year, size = 1, linetype = 2)+
  scale_x_continuous(limits = c(1900, 2016)) +
  coord_cartesian(expand = FALSE) +
  theme_classic() + 
  labs(title = "Age Distribution of American Boys Named Harper", 
       caption = "Source: Social Security Adminstration")
```

For boys, `r name_of_interest` peaked in `r peak$year` when `r max(df$n)` 
`r name_of_interest`s were born. Based on the actuarial data, 
414 male `r name_of_interest`'s from `r peak$year` 
would still be alive in `r lubridate::year(Sys.Date())`. Overall, 
`r round(sum(df$alive_count, na.rm=TRUE)/sum(df$n, na.rm=TRUE)*100,0)`% of a boy 
named `r name_of_interest` born between 1900 and 2017 would still be alive in 
`r lubridate::year(Sys.Date())`. The median age of a boy named 
`r name_of_interest` still alive, would be `r round(med_age$wted_median, 0)` years.

While the median age of Harpers is reasonably similar for both boys and girls
it's interesting to note that Harper only emerged as a girls name relatively 
recently. As such, if you know that someone is named Harper and born in 1955
you may feel confident guessing that they are more likely to be male than female.
^[Or rather more likely to be assigned male at birth] Whereas given the
relatively popularity of the names in recent years, a younger Harper is 
more likely to be female.^[Assigned female at birth]